Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-02-05 manuscript/manuscript.Rmd
substantial revision after reframing.

---
title: "Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology. More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill. It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].

@@

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

@@

The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project aims to examine the following research questions: 1) Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology? To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation. Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model during the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences; 3) Throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect. The third assumption here is oarticularly informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010; @preece2022]. In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)). As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019]. To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

@@

In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for expert knapper's (N. Khreisheh) review, leading to the compilation of a data set composing 9 assessments in total. It should be also noted that 6 out of 17 participants dropped out of the research before the final assessment due to personal reasons. To detect the effect of training on skill level and mental template, we reorganized our assessment classification scheme and combined it into three broader categories, namely pre-training (assessment 1), early training (assessment 2-5), and late training (assessment 6-9), which helps increase the sample size of the measured intervals. A more detailed experimental protocol can be assessed in one of our published papers [@pargeter2019].

@@

We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021]. As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of reduction intensity on morphological variation. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

In addition, visual inspection of the principle component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups. To test this, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)). Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.

@@

## Effects of training

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$). These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness). However, the effects of training across different assessment periods on both dimensions are not significant. Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant. It can also be inferred that the expert group display a higher variability in terms of delta weight compared with novices.

@@

Our study suggests that skill can differentially affect the expression of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction. In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2). This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill at Boxgrove, with potential implications for the cognitive demands and social contexts supporting learning [@pargeter2019; @stout2014]. It further suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investments in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation. In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill. These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error. Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to of a well-defined mental template [@garcía-medrano2019; @shipton2020].

Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, although some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

>>>>

---
title: "Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapping skill and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present new results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differential effect in the cultural reproduction of different aspects of handaxe morphology. More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill. Our findings thus shed new light on how the processes of skill learning can bias the cultural reproduction of artifact morphology. \\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Knapping skill; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019; @white2022]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapping skill [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].

@@

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

@@

The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The complexity of this issue is further exemplified by the fact that motor skills like knapping cannot be simply learned through observation but must be reconstructed through individual practice using supportive material in social contexts [@stout2017]. The ignorance of this factor becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. As we stated earlier, this reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of knapping skill and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project aims to examine the following research question:

> Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology? To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation. Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model before the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect different level of knapping skill; 3) Throughout the training the novice samples should become more similar to expert samples in both knapping skill and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect. The third assumption here is particularly informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020]. This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010; @preece2022]. In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high knapping skill level reflected in their manufacture (**Figure** \@ref(fig:photos)). As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019]. We selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017]. The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].

@@

In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas prepared by Bruce Bradley, which are part of our expert sample as described above. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for instructor's (N. Khreisheh) review, leading to the compilation of a data set composing 9 assessments in total. It should be also noted that 6 out of 17 participants dropped out of the research before the final assessment due to personal reasons. To understand the effect of skill acquisition on artifact morphology, we reorganized our assessment classification scheme and combined it into three broader categories, namely pre-training (assessment 1), early training (assessment 2-5), and late training (assessment 6-9), which helps increase the sample size of the measured intervals. A more detailed experimental protocol can be assessed in one of our published papers [@pargeter2019].

@@

We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021]. As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To understand the process of skill learning of novices using the Boxgrove and expert samples as benchmarks, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019], which makes it particularly suitable for this study as the sample size of expert experimental collection is rather small. Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of reduction intensity on morphological variation. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

In addition, visual inspection of the principle component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups. To test this intuition, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)). Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.

@@

## Effects of skill acquisition

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are not statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.65$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.33$, $p>0.05$). Additionally, the pairwise comparisons of mean PC2 values between the Early training and Expert ($t=-3.5$, $p>0.05$) and between the Late training and Expert ($t=-3.68$, $p>0.05$) are also not statistically significant. An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.82$, $p>0.05$). These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness), while the effects of training across different assessment periods on both dimensions are not significant. Interestingly, the post-training groups are very different from the expert group in the mean PC1 value, but not in the mean PC2 value. Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant. It can also be inferred that the expert group display a higher variability in terms of delta weight compared with novices.

@@

Our study suggests that skill can differentially affect the expression of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction. In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2). This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill at Boxgrove, with potential implications for the cognitive demands and social contexts supporting learning [@pargeter2019; @stout2014]. It further suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investments in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation. In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill. These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error. Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to a well-defined mental template [@garcía-medrano2019; @shipton2020].

Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating percussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies [@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, although some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

Unexpectedly, we found that modern experimental knappers did not closely approximate the PC2 (elongation and pointedness) values of Boxgrove handaxes. More specifically, the Boxgrove assemblage displays an ovate shape featuring a wider tip as previously pointed out by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019], while the experimental assemblages are characterized by a more pointed shape with a longer central axis (**Figure** \@ref(fig:GeneralPCA1)). This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar), but were not provided with a concrete template to copy. It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1) that likely reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182]. Novices then sought to approximate this form, as demonstrated by their instructor and exemplar handaxes from our current expert sample that were given as models, where the mean PC2 value of exemplar handaxes reach 1.18, indicating a high degree of elongation and pointedness even among the expert sample. It should be also noted that the instructor (N. Khreisheh) has learned how to knap and how to teach knapping from one of our expert knappers (B. Bradley), potentially suggesting a cascading effect of social learning that also contributed to a shared mental template between the expert group and the novice group after training. At a higher level, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide a mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise by actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.
